The dynamics of starvation and recovery

### 2016-08-02 Manuscript/6.tex
Kempes second round of edits

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features~\cite{}\sid{Ref}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $M$ is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans~\cite{}\sid{Ref}, and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. Several efforts have shown how a
partitioning of this metabolic rate between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{Kempes:2012hy,more}\sid{fix Ref}. More specifically, the cross-species trends in growth rate can be
approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development.  It is useful to explicitly write this balance
because it can also be modified to understand the timescales of starvation as
well as recovery from starvation as we show below.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$ (henceforth we term this state as the ``replete'' state), to a
reduced-mass hungry state where reproduction is impossible.  This transition
time can be inferred from the energy balance in Eq.~\ref{balance}, where we
make the basic assumption that an organism must meet its maintenance
requirements using the digestion of existing mass as the sole energy source.
This assumption implies a simple metabolic balance where the left hand side of Equation \ref{balance} becomes zero and we have \begin{eqnarray}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$~\cite{}\sid{Ref}, the energy required to
synthesis a unit of biomass. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-B_{m}t/E_{m}^{\prime}}.
\end{eqnarray}
Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation and
death by the timescales required to reach specific fractions of the
replete-state mass.  We define $m_{\rm starve}=\epsilon M$, where
$\epsilon<1$ is the fraction of replete-state mass where reproduction ceases.
This fraction will be modified if tissue composition systematically scales
with adult mass.  For example, making use of the observation that body fat in
mammals scales with overall body size according to $M_{f}=f_{0}M^{\gamma}$
and assuming that once this mass is fully digested the organism starves, this
would imply that $\epsilon=1-f_{0}M^{\gamma}/M$.  Using this criterion in
Eq.~\ref{mt}, the time scale for starvation is given by
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-\frac{E_{m}\ln\left(\epsilon\right)}{B_{m}}.
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Eq.~\ref{lambda}), which is important for the dynamics that we
later discuss.

The time to death should follow a similar relation, but defined by a lower
fraction of replete-state mass, $m_{\rm death}=\epsilon^{\prime} M$.
Suppose, for example, that an organism dies once it has digested all fat and
muscle tissues, and that muscle tissue scales with body mass according to
$M_{mm}=mm_{0}M^{\zeta}$ \sid{what is $mm_0$?}.  This gives
$\epsilon^{\prime}=1-\left(f_{0}M^{\gamma}+mm_{0}M^{\zeta}\right)/M$. Muscle
mass has been shown to be roughly proportional to body mass~\cite{muscle} in
mammals and thus $\epsilon^{\prime}$ is merely $\epsilon$ minus a constant.
Thus
\begin{eqnarray}
t_{\mu}=-\frac{E_{m}\ln\left(\epsilon^{\prime}\right)}{B_{m}}
\end{eqnarray}
and $\mu=1/t_{\mu}$.

@@

The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the hungry to the full state.  We again
use the balance given in Eq.~\ref{balance} to find the timescale for an
organism to return to the replete-state mass from a given reduced
hungry-state mass.  From the solution to Eq.~\ref{balance}
\begin{eqnarray}
m\left(t\right)\!=\!\left(\!\frac{B_{0}}{B_{m}}\!\right)^{1/(\eta-1)}\!\!
\left[1\!-\!\left(1\!-\!\frac{B_{m}}{B_{0}}m_{0}^{1\!-\!\eta}\right)e^{-b\left(1\!-\!\eta\right)t}\right]^{1/\left(1-\eta\right)}%\left(\frac{a}{b}\right)^{1/\left(1-\eta\right)}\,,
\end{eqnarray}
we require the timescale, $t_{\rho}=t_{2}-t_{1}$, which is the time it takes
to go from $m\left(t_{1}\right)=\epsilon M$ to $m\left(t_{2}\right)=M$, or
\begin{eqnarray}
t_{\rho}=\frac{E_{m}\left\{\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(M
  \right)^{1\!-\!\eta }\right]
-\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(\epsilon M \right)^{1\!-\!\eta }\right]\right\}}{(\eta -1) B_{m}}.
\end{eqnarray}
%\left(\frac{a}{b}\right)^{\frac{1}{\eta -1}}
%where $c=(a/b)^{1/(\eta-1)}$.
Although these rate equations are general, here we focus on parameterizations
for terrestrial-bound endotherms, specifically mammals, which range from a
minimum of $M\approx1$ gram (the Etruscan shrew \emph{Suncus etruscus}) to a
maximum of $M\approx10^7$ grams (the late Eocene to early Miocene
Indricotheriinae).  Investigating other classes of organisms would simply
involve altering the metabolic exponents and scalings associate with
$\epsilon$. Moreover, we emphasize that our allometric equations describe
mean relationships, and do not account for the (sometimes considerable)
variance associated with individual species.
\\
%, though this is beyond the current scope of our analysis. %, and we henceforth focus our examination on mammalian species.

@@

%Minimal mass bound
A lower bound on mammalian body size is given by $\epsilon=1$, where mammals
have no metabolic reserves and immediately starve; this occurs at a size of
\sid{$M=value$}.  This calculation \sid{what calculation?} gives an extreme
limit on size but does not account for the subtleties of starvation dynamics
that may limit body size. The NSM correctly predicts that species with
smaller masses have larger steady-state population densities.  However we
observe that there is a sharp change in the mass dependence of both the
steady-state densities and $\sigma/\rho$ at $M \approx 0.3$ grams
(Fig.~\ref{fig:mass}a,b).  The dependence of the rates of starvation and
recovery explain this phenomenon.  As the mass decreases, the rate of
starvation increases, while the rate of recovery declines super-exponentially
\sid{how do we know this?}.  This decline in $\rho$ occurs when the
percentage of body fat is
$1 - 1/\left[\big(B_{0}/B_{m}\big)^{1/\left(\eta-1\right)}M\right] \approx
2\%$,
whereupon consumers have no eligible route \sid{what does this mean?} that avoids
starvation.  Compellingly, this dynamic bound determined by the rate of
energetic recovery is close to the minimum observed mammalian body size of
ca. 1.3-2.5 grams (Fig.~\ref{fig:mass}b,c), a range that occurs as the
recovery rate begins its decline.  In addition to known transport
limitations~\cite{West:2002ud}, we suggest that an additional constraint of
lower body size stems from the dynamics of starvation.  This work \sid{which
  work?} mirrors other efforts where coincident limitations seem to limit the
smallest possibilities for life within a particular class or
organisms~\cite{kempes}.

@@

\begin{acknowledgments}
  C.P.K was supported by a Trump Fellowship from the American League of
  Conservatives.  S.R. was supported by grants DMR-1608211 and 1623243 from
  the National Science Foundation, and by the John Templeton Foundation.
\end{acknowledgments}

@@

% 
% 
%	 \begin{figure}[h]
%		 \centering
%		 \includegraphics[width=0.5\textwidth]{Growth-trajectory-diagram.pdf}
%		 \caption{
%		 }
%		 \label{growth-diagram}
%	 \end{figure}
%	 
%	 \begin{figure}[h]
%		 \centering
%		 \includegraphics[width=0.25\textwidth]{fig_GvS.pdf}
%		 \caption{
%		 }
%		 \label{GvS}
%	 \end{figure}
%
%
%	\begin{figure}[h]
%	 	\centering
%	 	\includegraphics[width=0.25\textwidth]{fig_FixedPoint.pdf}
%	 	\caption{
%	 	}
%		\label{Hopfb}
%	\end{figure}
%
%
%	\begin{figure}[h]
%		\centering
%		\includegraphics[width=0.8\textwidth]{fig_DataHopf.pdf}
%		\caption{
%		}
%		\label{DataHopf}
%	\end{figure}
%	 
% 
%	\begin{figure}[h]
%		\centering
%	 	\includegraphics[width=0.8\textwidth]{fig_ExtinctionAllometric.pdf}
%	 	\caption{
%	 	}
%	 	\label{Ext}
%	\end{figure}
%	
%	\begin{figure}[h]
%	 	\centering
%	 	\includegraphics[width=0.5\textwidth]{fig_FPAllometric.pdf}
%	 	\caption{
%	 	}
%	 	\label{Asymp}
%	\end{figure}
%
\clearpage

\begin{table}[h]
\caption{Parameter Values For Various Classes of Organisms}
\label{param}
    \begin{center}
    \small
     \begin{tabular}{| p{1.2cm}| p{3.2cm} | p{2.6cm} | p{3.2cm} | }
     \hline
     & {\bf Mammals} & {\bf Unicellular Eukaryotes} & {\bf Bacteria} \\
     \hline
   $\eta$ & $3/4$ & & $1.70$ \\ 
   $E_{m}$ & $10695$ (J gram$^{-1}$) & & $10695$ (J gram$^{-1}$) \\ 
   $E_{m}^{\prime}$ & $\approx E_{m}$ & & $\approx E_{m}$ \\ 
   $B_{0}$ & $0.019$ (W gram$^{-\alpha}$) & & $1.96\times10^{17}$ \\
   $B_{m}$ & $0.025$ (W gram$^{-1}$)   & & $0.025$ (W gram$^{-1}$)\\
   $a$ & $1.78\times10^{-6}$ & & $1.83\times10^{13}$ \\ 
   $b$ & $2.29\times10^{-6}$ & & $2.29\times10^{-6}$ \\  
   $\eta-1$ & $-0.21$ & & $0.73$ \\ 
   $\lambda_{0}$ & $3.39\times10^{-7}$ (s$^{-1}$ gram$^{1-\eta}$) & & $56493$ \\ 
   $\gamma$ & $1.19$ & & $0.68$ \\ 
   $f_{0}$ & $0.02$ & & $1.30\times10^{-5}$ \\ 
   $\zeta$ & $1.01$ & & \\ 
   $mm_{0}$ & $0.32$ & & \\

   \hline
    \end{tabular}
    \end{center}
   \end{table}

>>>>

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features \cite{Brown:2004wq}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $M$ is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans (e.g. \cite{Brown:2004wq}), and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. Several efforts have shown how a
partitioning of this metabolic rate between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}. More specifically, the cross-species trends in growth rate can be
approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development.  This balance has the general solution \cite{moses2008rmo, Kempes:2012hy}
\begin{eqnarray}
m\left(t\right)\!=\left[1\!-\!\left(1\!-\!\frac{b}{a}m_{0}^{1\!-\!\eta}\right)e^{-b\left(1\!-\!\eta\right)t}\right]^{1/\left(1-\eta\right)}\!\left(\!\frac{a}{b}\!\right)^{1/(1-\eta)}
\end{eqnarray}
where, $a=B_{0}/E_{m}$ and $b=B_{m}/E_{m}$ , andwhich we use to define the timescale of recovery from starvation. The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the hungry to the full state.  Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation,
death, and recovery by the timescales required to reach, or return from, specific fractions of the
replete-state mass. We define $m_{\rm starve}=\epsilon M$, where
$\epsilon<1$ is the fraction of replete-state mass where reproduction ceases. This fraction will be modified if tissue composition systematically scales
with adult mass.  For example, making use of the observation that body fat in
mammals scales with overall body size according to $M_{f}=f_{0}M^{\gamma}$
and assuming that once this mass is fully digested the organism starves, this
would imply that $\epsilon=1-f_{0}M^{\gamma}/M$. It follows that the recovery timescale, $t_{\rho}=t_{2}-t_{1}$, is the time it takes
to go from $m\left(t_{1}\right)=\epsilon M$ to $m\left(t_{2}\right)=M$, or
\begin{eqnarray}
t_{\rho}=\frac{\ln\left[1\!-\!\frac{a}{b}\left(M
  \right)^{1\!-\!\eta }\right]
-\ln \left[1\!-\!\frac{a}{b}\left(\epsilon M \right)^{1\!-\!\eta }\right]}{(\eta -1) b}.
\end{eqnarray}
It should be noted that more complicated ontogenetic models explicitly handle storage \cite{hou}, whereas this is implicitly covered by the body fat scaling in our framework.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$ (henceforth we term this state as the ``replete'' state), to a
reduced-mass hungry state where reproduction is impossible.  For starving individuals we assume that an organism must meet its maintenance requirements using the digestion of existing mass as the sole energy source.
This assumption implies the following simple metabolic balance 
\begin{eqnarray}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$, the energy required to
synthesis a unit of biomass \cite{hou}. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-b^{\prime}t}
\end{eqnarray}
where $b^{\prime}=E_{m}^{\prime}/B_{m}$. The time time scale for starvation is given by the time it takes this decay to reach $\epsilon M$, which is
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-b^{\prime}\ln\left(\epsilon\right).
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Eq.~\ref{lambda}), which is important for the dynamics that we
later discuss.

The time to death should follow a similar relation, but defined by a lower
fraction of replete-state mass, $m_{\rm death}=\epsilon^{\prime} M$.
Suppose, for example, that an organism dies once it has digested all fat and
muscle tissues, and that muscle tissue scales with body mass according to
$M_{m}=u_{0}M^{\zeta}$.  This gives
$\epsilon^{\prime}=1-\left(f_{0}M^{\gamma}+u_{0}M^{\zeta}\right)/M$. Muscle
mass has been shown to be roughly proportional to body mass~\cite{muscle} in
mammals and thus $\epsilon^{\prime}$ is merely $\epsilon$ minus a constant.
Thus
\begin{eqnarray}
t_{\mu}=-b^{\prime}\ln\left(\epsilon^{\prime}\right)
\end{eqnarray}
and $\mu=1/t_{\mu}$.

@@

%\left(\frac{a}{b}\right)^{\frac{1}{\eta -1}}
%where $c=(a/b)^{1/(\eta-1)}$.
Although these rate equations are general, here we focus on parameterizations
for terrestrial-bound endotherms, specifically mammals, which range from a
minimum of $M\approx1$ gram (the Etruscan shrew \emph{Suncus etruscus}) to a
maximum of $M\approx10^7$ grams (the late Eocene to early Miocene
Indricotheriinae).  Investigating other classes of organisms would simply
involve altering the metabolic exponents and scalings associate with
$\epsilon$. Moreover, we emphasize that our allometric equations describe
mean relationships, and do not account for the (sometimes considerable)
variance associated with individual species.
\\
%, though this is beyond the current scope of our analysis. %, and we henceforth focus our examination on mammalian species.

@@

%Minimal mass bound
A lower bound on mammalian body size is given by $\epsilon=1$, where mammals
have no metabolic reserves and immediately starve; this occurs at a size of
\sid{$M=value$}.  This calculation \sid{what calculation?} gives an extreme
limit on size but does not account for the subtleties of starvation dynamics
that may limit body size. The NSM correctly predicts that species with
smaller masses have larger steady-state population densities.  However we
observe that there is a sharp change in the mass dependence of both the
steady-state densities and $\sigma/\rho$ at $M \approx 0.3$ grams
(Fig.~\ref{fig:mass}a,b).  The dependence of the rates of starvation and
recovery explain this phenomenon.  As the mass decreases, the rate of
starvation increases, while the rate of recovery declines super-exponentially
\sid{how do we know this?}.  This decline in $\rho$ occurs when the
percentage of body fat is
$1 - 1/\left[\big(a/b\big)^{1/\left(\eta-1\right)}M\right] \approx
2\%$,
whereupon consumers have no eligible route \sid{what does this mean?} that avoids
starvation.  Compellingly, this dynamic bound determined by the rate of
energetic recovery is close to the minimum observed mammalian body size of
ca. 1.3-2.5 grams (Fig.~\ref{fig:mass}b,c), a range that occurs as the
recovery rate begins its decline.  In addition to known transport
limitations~\cite{West:2002ud}, we suggest that an additional constraint of
lower body size stems from the dynamics of starvation.  This work \sid{which
  work?} mirrors other efforts where coincident limitations seem to limit the
smallest possibilities for life within a particular class or
organisms~\cite{Kempes:2012hy}.

@@

\begin{acknowledgments}
  J.D.Y. was supported by an Omidyar postdoctoral Fellowship at the Santa Fe Institute. 
  C.P.K was supported by a Trump Fellowship from the American League of
  Conservatives.  
  S.R. was supported by grants DMR-1608211 and 1623243 from
  the National Science Foundation, and by the John Templeton Foundation.
\end{acknowledgments}

@@

% 
% 
%	 \begin{figure}[h]
%		 \centering
%		 \includegraphics[width=0.5\textwidth]{Growth-trajectory-diagram.pdf}
%		 \caption{
%		 }
%		 \label{growth-diagram}
%	 \end{figure}
%	 
%	 \begin{figure}[h]
%		 \centering
%		 \includegraphics[width=0.25\textwidth]{fig_GvS.pdf}
%		 \caption{
%		 }
%		 \label{GvS}
%	 \end{figure}
%
%
%	\begin{figure}[h]
%	 	\centering
%	 	\includegraphics[width=0.25\textwidth]{fig_FixedPoint.pdf}
%	 	\caption{
%	 	}
%		\label{Hopfb}
%	\end{figure}
%
%
%	\begin{figure}[h]
%		\centering
%		\includegraphics[width=0.8\textwidth]{fig_DataHopf.pdf}
%		\caption{
%		}
%		\label{DataHopf}
%	\end{figure}
%	 
% 
%	\begin{figure}[h]
%		\centering
%	 	\includegraphics[width=0.8\textwidth]{fig_ExtinctionAllometric.pdf}
%	 	\caption{
%	 	}
%	 	\label{Ext}
%	\end{figure}
%	
%	\begin{figure}[h]
%	 	\centering
%	 	\includegraphics[width=0.5\textwidth]{fig_FPAllometric.pdf}
%	 	\caption{
%	 	}
%	 	\label{Asymp}
%	\end{figure}
%
%\clearpage

%\begin{table}[h]
%\caption{Parameter Values For Various Classes of Organisms}
%\label{param}
%    \begin{center}
%    \small
%     \begin{tabular}{| p{1.2cm}| p{3.2cm} | p{2.6cm} | p{3.2cm} | }
%     \hline
%     & {\bf Mammals} & {\bf Unicellular Eukaryotes} & {\bf Bacteria} \\
%     \hline
%   $\eta$ & $3/4$ & & $1.70$ \\ 
%   $E_{m}$ & $10695$ (J gram$^{-1}$) & & $10695$ (J gram$^{-1}$) \\ 
%   $E_{m}^{\prime}$ & $\approx E_{m}$ & & $\approx E_{m}$ \\ 
%   $B_{0}$ & $0.019$ (W gram$^{-\alpha}$) & & $1.96\times10^{17}$ \\
%   $B_{m}$ & $0.025$ (W gram$^{-1}$)   & & $0.025$ (W gram$^{-1}$)\\
%   $a$ & $1.78\times10^{-6}$ & & $1.83\times10^{13}$ \\ 
%   $b$ & $2.29\times10^{-6}$ & & $2.29\times10^{-6}$ \\  
%   $\eta-1$ & $-0.21$ & & $0.73$ \\ 
%   $\lambda_{0}$ & $3.39\times10^{-7}$ (s$^{-1}$ gram$^{1-\eta}$) & & $56493$ \\ 
%   $\gamma$ & $1.19$ & & $0.68$ \\ 
%   $f_{0}$ & $0.02$ & & $1.30\times10^{-5}$ \\ 
%   $\zeta$ & $1.01$ & & \\ 
%   $mm_{0}$ & $0.32$ & & \\ 
%   
%      
%   \hline
%    \end{tabular}
%    \end{center}
%   \end{table}
